Forcing for Isabelle

### 2018-12-20 LICS/intro.tex
una parte de la intro. Faltan algunas cosas

One of the main initial concerns was to decide if this foundation
completely axiomatized all of mathematics. As it was shown by G\"odel
in 1931, no  such axiomatization (therefore, including artihmetic) can
be complete. A refinement of this result, is G\"odel's Second
Incompleteness Theorem that says that no consistent formal system that
includes arithmetic can prove its own consistency. As a matter of
fact, many problem within the area of Set Theory, not directly related
to consistency or logic, have been proved to be undecided by the
theory. Two techniques have been developed to do this. One is G\"odel
theory of \emph{inner models}, which started with his model $L$ of
\emph{Axiom of Constructibility} and was used to proved the relative
consistency of the Axiom of Choice and the Generalized Continuum
Hypothesis with $\ZF$.  he has been formalized by Lawrence
Paulson\cite{paulson_2003} in the proof assistant \emph{Isabelle},

@@

An enumeration of achievements to be presented:
\begin{enumerate}
\item Discussion of the Isabelle view of life.
  \begin{enumerate}
  \item Meta (Pure) vs Object (FOL+ZF) theory.
  \item Object theory is presented axiomatically; in particular,
    first-order ``formulas'' are not such (i.e., the \emph{initial}
    algebra of formulas is \textbf{not} being specified). Thus, we are
    not able to perform recursion on terms of type \tyo.
  \end{enumerate}
\item Discussion of locales.
  \begin{enumerate}
  \item How they help presenting complex results.
  \item Their key role in stating the Fundamental Theorems (without
    defining \isatt{forces}).
  \item \isatt{forces} is defined for elements of \formula.
  \end{enumerate}
\item Enhancements in the recursion machinery.
  \begin{enumerate}
  \item Impact on names.
  \end{enumerate}
\item Hacking of \isatt{ZF-Constructible}.
  \begin{enumerate}
  \item Elimination of the \tyo-schematic Replacement Axiom from
    locales.
  \item Replacing \isatt{M(nat)} by \isatt{M(0)} (merely non
    emptiness) from Relative (and other places?).
  \end{enumerate}
\item Inteface between first-order properties of $M$ and those of
  $\#\#M$.
  \begin{enumerate}
  \item Enumerate properties and statistics.
  \end{enumerate}
\item Details of the proof of Separation.
\end{enumerate}

>>>>

One of the main initial concerns was to decide if this foundation
completely axiomatized all of mathematics. As it was shown by G\"odel
in 1931, no  such axiomatization (therefore, including artihmetic) can
be complete. A refinement of this result, is G\"odel's Second
Incompleteness Theorem that says that no consistent formal system that
includes arithmetic can prove its own consistency. As a matter of
fact, many problem within the area of Set Theory, not directly related
to consistency or logic, have been proved to be undecided by the
theory. Two techniques have been developed to do this. One is Kurt G\"odel's
theory of \emph{inner models}, which started with his model $L$ of
\emph{Axiom of Constructibility} \cite{godel-L} and was used to proved the relative
consistency of the Axiom of Choice and the Generalized Continuum
Hypothesis with $\ZF$. The second one is the technique of
\emph{forcing}, which was developed 30 years later by Paul
J. Cohen~\cite{Cohen-CH-PNAS}.

The constructible universe  has been formalized by Lawrence
Paulson \cite{paulson_2003} in the proof assistant \emph{Isabelle}, and
the present paper is part of a project of formalizing the whole of the
forcing machinery, continuing the work by Paulson. The starting point
of this project is our paper \cite{2018arXiv180705174G}, to which we
refer the reader for further details on basic matters.

@@

We briefly describe the contents of each
section. Section~\ref{sec:isabelle} contains the bare minimun
requirements to understand the (meta)logics used in Isabelle. Next, an
overview of the model theory of set theory is presented in
Section~\ref{sec:axioms-models-set-theory}. There is an ``internal''
representation of first-order formulas as sets, implemented by
Paulson; Section~\ref{sec:renaming} discusses syntactical
transformations of the former, mainly permutation of variables. 
In Section~\ref{sec:generic-extensions} the generic extensions are
succintly reviewed and how the treatment of well founded recursion in
Isabelle was enhanced. We take care of the ``easy axioms'' in
Section~\ref{sec:easy-axioms}; these are the ones that
do not depend on the forcing theorems. We describe the latter in
Section~\ref{sec:forcing}. We adapted the  work by Paulson to our
needs, and this is described in
Section~\ref{sec:hack-constructible}. We present the proof
of the Separation Axiom Scheme in Section~\ref{sec:proof-separation},
which follows closely its implementation. A plan for future work and
some immediate conclusions are offered in
Section~\ref{sec:conclusions-future-work}.

An enumeration of achievements to be presented:
\begin{enumerate}
\item Discussion of locales.
  \begin{enumerate}
  \item How they help presenting complex results.
  \item Their key role in stating the Fundamental Theorems (without
    defining \isatt{forces}).
  \item \isatt{forces} is defined for elements of \formula.
  \end{enumerate}
\item Enhancements in the recursion machinery.
  \begin{enumerate}
  \item Impact on names.
  \end{enumerate}
\item Inteface between first-order properties of $M$ and those of
  $\#\#M$.
\end{enumerate}
